---
title: "modeling"
output: html_document
date: "2025-03-25"
---

```{r setup, include=FALSE}
#pacman to check whether packages are installed, if not load them
if (!require("pacman")) install.packages("pacman")
library(pacman)
pacman::p_load(dplyr,
 tidyverse,
 tidymodels,
 purrr,
 ggplot2,
 gridExtra,
 MASS,
 readr,
 psych,
 caret,
 class,
 rpart,
 rattle
)
```

## Data Introduction

The [Taiwanese Bankruptcy Prediction dataset](https://archive.ics.uci.edu/dataset/572/taiwanese+bankruptcy+prediction) was donated on 6/27/2020. The data were collected from the Taiwan Economic Journal, years 1999-2009. They defined company bankruptcy based on the business regulations of the Taiwan Stock Exchange. This is a multivariate dataset with a classification outcome (Bankrupt or not). There are a total of 6,819 observations and 95 variables. We are interested in a subset of variables to predict bankruptcy:

-   Operating Gross Margin (continuous)

-   Operation Profit Rate (continuous)

-   Operating Express Rate (continuous)

-   Revenue per person (continuous)

-   Degree of Financial leverage (DFL) (continuous)

Before any analysis of the data, we will first perform exploratory data analysis.

## Loading in the data

```{r data import}
bank_data <- read_csv("data.csv")

#filtering down to columns of interest 
bank_data <- bank_data %>% dplyr::select("Bankrupt?","Operating Gross Margin","Operating Profit Rate","Operating Expense Rate","Revenue per person","Degree of Financial Leverage (DFL)")

#renaming variables for easier column selection
colnames(bank_data) = c("Bankrupt","OGM","OPR","OER","RPP","DFL")

#changing bankrupt to a factor variable 
bank_data$Bankrupt = as.factor(bank_data$Bankrupt)

describe(bank_data)
```

There are no missing variables. All summary statistics seem reasonable for each data type.

Checking the correlation between variables:

```{r corr_check}
cor(bank_data[c("OGM","OPR","OER","RPP","DFL")])
```

There does not appear to be any of-note correlation/linear relationships between variables. The strongest correlation is between operating gross margin and operating expense rate at -0.21, which is still a week correlation.

We next want to check if assumptions of normality are appropriate. First, we can look at univariate normality via QQ plots.

```{r}
# Create a list of columns to plot
data_list <- bank_data[, c("OGM", "OPR", "OER", "RPP", "DFL")]

# Generate QQ plots for each column
qq_plots <- lapply(names(data_list), function(name) {
  ggplot(data.frame(x = data_list[[name]]), aes(sample = x)) +
    stat_qq() + stat_qq_line() +
    ggtitle(paste("QQ Plot of", name)) +
    theme_minimal()
})

# Arrange all plots in a grid
grid.arrange(grobs = qq_plots, nrow = 2, ncol = 3)
```

For OGM, OPR, RPP, and DFL, there are a few minor departures from normality, but considering how large our data set is, we can assume normality for these variable.

We see large departures from normality for the OER variable. The Shapiro-Wilk normality test could be used, however this test is for samples of under 5000 size. Because we are looking at data of very large size, we should be okay to assume univariate normality for all variables.

Central Limit Theorem (CLT) says when n is large enough, X\^ approximately has a N(u,sigma\^2/n) distribution. This holds for multivariate normal distribution.

## Modeling 

train test split

```{r}
set.seed(123) #for reproducibility

#80/20 train test split 
train_index <- initial_split(bank_data, prop = 0.8)
train_data <- training(train_index)
test_data <- testing(train_index)
```

set up cv

```{r}
ctrl <- trainControl(method = "cv", number = 10, classProbs = TRUE, summaryFunction = twoClassSummary, savePredictions = "final")

```

kNN model (tuning K)

**Tuned Parameter:** `k`: Number of neighbors considered when making a prediction

```{r}
# Convert outcome to factor with level names
train_data$`Bankrupt` <- factor(train_data$`Bankrupt`, labels = c("No", "Yes"))
test_data$`Bankrupt` <- factor(test_data$`Bankrupt`, labels = c("No", "Yes"))

knn_fit <- train(
  `Bankrupt` ~ ., data = train_data,
  method = "knn",
  trControl = ctrl,
  tuneLength = 10,  # automatically tries 10 values of k
  metric = "ROC"
)
```

logistic regression

```{r}
logit_fit <- glm(`Bankrupt` ~ ., data = train_data, family = binomial)

# Predict on test set
logit_probs <- predict(logit_fit, newdata = test_data, type = "response")
logit_preds <- ifelse(logit_probs > 0.5, "Yes", "No")
logit_preds <- factor(logit_preds, levels = c("No", "Yes"))

logit_fit$coefficients
```

classification tree

**Tuned Parameter:**

`cp`: Complexity parameter (controls pruning --- higher `cp` means smaller tree)

```{r}

tree_fit <- train(
  `Bankrupt` ~ ., data = train_data,
  method = "rpart",
  trControl = ctrl,
  tuneLength = 10
)


# Visualize the tree
fancyRpartPlot(tree_fit$finalModel)
```

evaluate models

```{r}
# For caret models
eval_model <- function(model, test_data) {
  preds <- predict(model, newdata = test_data)
  cm <- confusionMatrix(preds, test_data$`Bankrupt`)
  aper <- 1 - cm$overall["Accuracy"]
  list(Accuracy = cm$overall["Accuracy"], Kappa = cm$overall["Kappa"], APER = aper)
}

knn_eval <- eval_model(knn_fit, test_data)
tree_eval <- eval_model(tree_fit, test_data)

# For logistic regression (separate since it's not caret)
logit_cm <- confusionMatrix(logit_preds, test_data$`Bankrupt`)
logit_aper <- 1 - logit_cm$overall["Accuracy"]
logit_eval <- list(Accuracy = logit_cm$overall["Accuracy"], Kappa = logit_cm$overall["Kappa"], APER = logit_aper)

```

compare models

```{r}
model_results <- data.frame(
  Model = c("kNN", "Logistic Regression", "Decision Tree"),
  Accuracy = c(knn_eval$Accuracy, logit_eval$Accuracy, tree_eval$Accuracy),
  Kappa = c(knn_eval$Kappa, logit_eval$Kappa, tree_eval$Kappa),
  APER = c(knn_eval$APER, logit_eval$APER, tree_eval$APER)
)

print(model_results)
```
